Reproducibility of SNP-calling in multiple sequencing runs from single tumors

### 2015-05-18 PAPER/paper.tex
Intro and first part of Results

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{lscape}
% These packages are all incorporated in the memoir class to one degree or another...

@@

We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling when using the same data analysis pipeline on two instances of sequencing the same tumor, and to determine what additional filtering might increase or decrease the total similarity. We analyzed 55 pairs of technical replicates using the same computational pipeline and measured the extent of the overlap between two replicates, that is how many specific point mutations were found in both replicates. We further tested whether additional filtering increased or decreased the size of the overlap.  We found that about half of the putative mutations found in one sequencing run of a given sample are also found in the second, and that this percentage remains steady throughout orders of magnitude of variation in the total number of mutations called (23 to 10,966). We further found that using a SNP and subsequent filtering removes the overlap completely. We concluded that there is variation in the frequency of mutations in GBMs, and that while some parts of the SNP-caller preferentially removed putative mutations found in only one replicate, others removed primarily putative mutations found in both.

@@

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from a single patient for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}. TCGA's database includes hundreds of samples of each of several tumor types, including 528 and counting samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival of 14 months and a 5-year survival of ~5\%, and prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}. To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large scale genomic data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.

But these and other efforts have produced very few results for patients. It may be that we simply need more new ways to analyze the data. Another factor, however, may be that the data are noisy. It is well known that sequencing and SNP-calling pipelines are not error-free. Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal to noise ratio in this data set may be even worse. Despite widespread use of the TCGA dataset, and significant monetary investment in collecting and analyzing the data, we found very little research concerning the best way to maximize the quality of the sequence data and SNP-calls. Because different SNP-calling pipelines have been known to return different results on the same data \citep{SNPcall}, we turned this question around and asked, will the same SNP-calling pipeline return the same results on two sequencing runs from the same tissue? Does SNP-calling software and filtering increase or decrease the degree similarity?

We answered these questions using data from TCGA, which includes 55 GBM tumors that were sequenced twice, once each with (i) the standard WGS protocol, and (ii) an additional amplification step before library prep, which we refer to as the WGA protocol. For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \citep{SomaticSniper}. We found significant overlap (around 50\%) between technical replicates, but also significant differences. As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates. Still, the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192. Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of mutations, suggesting real variation in mutation frequency between samples. This may be in part due to know mutational hotspots in some tumors \citep{Karen}. Our attempt to use SomaticSniper and some custom filters to increase the similarity between technical replicates was unsuccessful: filtering removed almost the entirety of the overlap between the replicates.

@@

\subsection*{How similar are two iterations of sequencing the same tumor?}

@@

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis \citep{seqerror}. Loss of heterozygosity also occurs in sequencing if only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: a mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or not at all. In general, we would like to know how often these error occur. One way of investigating would be to use PCR to verify each individual mutation, but this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times, and look at the similarity between replicates. Theoretically, any fully penetrant mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not. (Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) This is the method used in most biological sequencing experiments--but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Still, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know about what percentage of called mutations would or would not likely appear in additional sequencing replicates.

TCGA's GBM data set includes 2 technical replicates each of 55 tumors. In this case, the technical replicates are not identical. One protocol included additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA replicate, and the other as the WGS replicate. Despite this difference, we still assumed that any fully penetrant mutation would appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one. Thus overall, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate. We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNPs per sample, than the corresponding WGS samples.

For each patient (n=55), we called mutations in both technical replicates and the patient's blood sample using the same computational pipeline (see Figure 1 and Methods): we downloaded TCGA bamfiles with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}. We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called in each replicate and the number of individual somatic mutations called in both replicates (hereafter, the length of the overlap, see Figure 2). We further calculated the percentage of of mutations in each replicate that occurred in the overlap between the two.

There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (for more detail, see summary statistics, Table 1). Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure 2). As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures 3 and 4). We further found that the percent overlap between the two samples, calculated by $WGA \cap WGS/WGS$ for WGS replicates and $WGA \cap WGS/WGA$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (see summary statistics, Table 1). As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure 3).

% Experiment: number of mutations (figure 4)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 23 to over 10,966). Different cancers mutate at different rates: some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly more mutations, usually resulting from errors in DNA repair pathways \citep{mutator}. GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome). One possible explanation is a degraded DNA sample, or bad data. If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations. We found no correlation (cor=0.25, p=.06, not significant) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure 4), though it is possible that with more more samples this could become a weak correlation. Still, that the percentage overlap does not correlate strongly with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.

@@

We next looked at the individual effects of each of the five filters that we coded and one of the SomaticSniper filters (Variant Allele Quality, or VAQ). We first looked at the total number of putative SNPs removed by each filter (Figure 5). We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 and 539 putative SNPs, respectively (see Table 1 for summary statistics). Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample (Figure 5, summary statistics in Table 1). The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average.

We next asked how many of the mutations removed by each filter were mutations present in the overlap between technical replicates, and how many were in just one sample? Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only one replicate)? To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure 6). We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%, 4\%, or 3\% of the overlap, or almost none, respectively (summary statistics in Table 1). In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, or about half each, respectively (Figure 6, summary statistics in Table 1). Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.

%% this is figures 8 and 9, which need to be added to the document and explained now...
Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different putative SNPs in each sample. We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (Figure 7), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (Figure 8). We found opposite trends: the percentage filtered out by VAQ decreased, and the percentage filtered out by LOH increased, with increasing length of overlap. The two graphs are almost, but not quite, perfect inverses. But given that (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time, this is expected. We concluded that the LOH and VAQ filters were not removing the same putative mutations.

@@

All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}. We downloaded three bamfiles for each of 55 patients using CGHub \citep{CGHub}. For each patient, data consisted of one bamfile taken from blood DNA, and two bamfiles from tumor DNA, one for each technical replicate. In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.

We created a pipeline for backend processing of all TCGA bamfiles, summarized in Figure 1, with commands given in Table 3. The custom python code to connect the pipeline is available in a public 
git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}). Our pipeline first regenerated fastq files (original reads) from the TCGA bamfiles, which were aligned to hg18, using picard \citep{picard}. Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new bamfile. We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}. Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.

@@

After generating a VCF file will all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates. We then filtered the lists of putative SNPs according to the eight filters described in Table 1, using a combination of command line options and custom python code (available in a public git repository). The eight filters were enacted as follows:

\begin{enumerate}
\item (GATK) The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2). We did not at any point consider putative mutations removed by this filter, and did not consider its individual action. 
\item (SS) The SomaticScore is a similar metric calculated by SomaticSniper. As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2). We did not at any point consider putative mutations removed by this filter, and did not consider its individual action. 
\item (VAQ) The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type. SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item (LOH) It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors. We excluded LOH variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item (10bp-SNP) It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment. We excluded 10bp-SNP variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item (10bp-INDEL) It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment. We excluded 10bp-INDEL variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item (dbSNP) It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant. We excluded dbSNP variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item ($<10\%$) It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error. We excluded $<10\%$ variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\end{enumerate}

@@

%% Summary Statistics Table
\begin{landscape}
\begin{table}
\caption{\textbf{TABLE CAPTION} Table explanation.}
\begin{tabular}{ p{8cm} p{2cm} p{2cm} p{2cm}p{2cm}p{2cm} }
	Quantity & average & median & min & max & stdev \\
	\hline
	No. mutations, WGS & 844 & 328 & 110 & 8,192 & 1,306 \\
	No. mutations, WGA & 1,531 & 694 & 23 & 10,966 & 2,230 \\
	\% mutations in overlap, WGS & 31\% & 31\% & 1\% & 74\% & 20\% \\
	\% mutations in overlap, WGA & 44\% & 45\% & 3\% & 71\% & 13\% \\
	No. putative mutations removed, VAQ & 309 & 164 & 17 & 11,439 & 1,372 \\
	No. putative mutations removed, LOH & 539 & 169 & 3 & 8,538 & 1,051 \\
	No. putative mutations removed, 10bp-SNP & 46 & 19 & 0 & 1,515 & 115 \\
	No. putative mutations removed, 10bp-INDEL & 28 & 16 & 0 & 535 & 45 \\
	No. putative mutations removed, dbSNP & 16 & 8 & 0 & 348 & 33 \\
	No. putative mutations removed, $<$10\% & 1 & 1 & 1 & 36 & 3 \\
	\% overlap removed, VAQ & 53\% & 57\% & 0\% & 100\% & 28\% \\
	\% overlap removed, LOH & 51\% & 52\% & 0\% & 99\% & 29\% \\
	\% overlap removed, 10bp-SNP & 3\% & 2\% & 0\% & 20\% & 3\% \\
	\% overlap removed, 10bp-INDEL & 4\% & 4\% & 0\% & 22\% & 4\% \\
	\% overlap removed, dbSNP & 3\% & 2\% & 0\% & 47\% & 7\% \\
\end{tabular}
\end{table}
\end{landscape}

>>>>

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
\usepackage{lscape}
\usepackage{amsmath}
% These packages are all incorporated in the memoir class to one degree or another...

@@

We examined 55 technical sequencing replicates of Glioblastoma multiforme (GBM) tumors from The Cancer Genome Atlas (TCGA) to ascertain the degree of repeatability in SNP-calling. We used the same computational pipeline on all pairs of samples, and we measured the extent of the overlap between two replicates, that is, how many specific point mutations were found in both replicates. We further tested whether additional filtering increased or decreased the size of the overlap.  We found that about half of the putative mutations identified in one sequencing run of a given sample were also identified in the second, and that this percentage remained steady throughout orders of magnitude of variation in the total number of mutations identified (from 23 to 10,966). We further found that using filtering after SNP-calling removed the overlap completely. We concluded that there is variation in the frequency of mutations in GBMs, and that while some parts of the SNP-caller preferentially removed putative mutations found in only one replicate, others removed primarily putative mutations found in both.

@@

The past six years have seen an explosion of data in cancer genomics, an effort led by TCGA, an archive of publicly-available data that includes sequencing of paired tumor-normal samples from individual patients for thousands of tumors \citep{TCGA-GBM, TCGA-GBM-13}. TCGA's database includes hundreds of samples of each of several tumor types, including over 500 samples of Glioblastoma multiforme (GBM), the most common and deadly primary brain tumor. GBM has a median survival time of 14 months and a 5-year survival rate of ~5\%. Prognosis for patients with this disease remains poor despite significant research investment, due to the difficulty of surgical resection and the limited number of effective chemotherapeutics \citep{GBM-stats}. To work towards the goal of improving patient outcomes using precision medicine, TCGA and other groups \citep{Parsons} used large scale genomic data to discover genes and pathways mutated in GBM \citep{pathways}, discover different GBM subtypes \citep{subtypes}, and develop a variety of computational models to find GBM driver mutations \citep{drivers}.

However, these and other efforts have produced very few results for patients in terms of improved treatments. It may be that we simply haven't yet analyzed the data in the most informative manner. Another factor may be that the data are noisy. It is well known that sequencing and SNP-calling pipelines are not error-free. For example, different SNP-calling pipelines can return different results on the same data \citep{SNPcall}. Given the heterogeneity in cancer genomes \citep{heterogenous1, heterogenous2}, and the presence of functional low-frequency variants in GBM \citep{rare}, the signal-to-noise ratio in the TCGA dataset may be particularly low. Yet despite widespread use of this dataset, and significant monetary investment in collecting and analyzing the data, we know little about how to maximize the quality of the sequence data and SNP-calls. Here, we addressed one aspect of this question, by asking whether the same SNP-calling pipeline will return the comparable results on two sequencing runs from the same tissue. And further, we asked whether added filtering after SNP calling increases or decreases the degree of similarity.

We answered these questions using TCGA data from 55 GBM tumors that were sequenced twice, once each with (i) the standard whole-genome sequencing (WGS) protocol, and (ii) an additional amplification step before library prep, which we refer to as the whole-genome amplification (WGA) protocol. For each of these 55 technical replicate pairs, we compared the somatic variants found in the WGS and WGA replicates, before and after analyzing the variants using SomaticSniper \citep{SomaticSniper}. We found significant overlap (around 50\%) between technical replicates, but also significant differences. As expected, the additional amplification step in the WGA protocol versus the WGS protocol added some putative mutations to the sample, so that on average these replicates had (i) more putative mutations, and (ii) a smaller percentage overlap between replicates. We found that the number of mutations in the WGS replicates varied by orders of magnitude, from 110 to 8,192. Contrary to expectations, the percentage overlap between technical replicates did not decrease with increasing numbers of putative mutations, suggesting real variation in mutation frequency between samples. This result may be in part due to known mutational hotspots in some tumors \citep{Karen}. Our attempt to use SomaticSniper filters, as well as some custom filters, to increase the similarity between technical replicates was unsuccessful: Filtering removed almost the entirety of the overlap between the replicates.

@@

\subsection*{How similar are two replicate sequencing runs of the same tumor?}

@@

DNA sequencing is not error-free. Error is introduced by mis-called bases in sequencing runs, and by mis-aligned bases during sequence analysis \citep{seqerror}. Loss of heterozygosity also occurs in sequencing if only one allele of a polymorphic site is amplified. Cancer DNA is highly heterogenous, which makes for an additional source of error: A mutation present in only some tumor cells may or may not be present in a given sample at high enough frequency to be seen, so that a partially penetrant mutation may appear fully penetrant or completely absent. In general, we would like to know how often these errors occur. One way of investigating this question would be to use PCR to verify each individual mutation. However, this method is expensive and time consuming. A cheaper alternative would be to sequence the tumor multiple times and to look at the similarity between replicates. Theoretically, any fully penetrant mutation will appear in all replicates, while errors due to (i) sequencing errors, (ii) amplification errors, or (iii) alignment errors will not. (Incompletely penetrant mutations would be present in some but not all samples, so this method does not address the difficulty of calling low-frequency variants in tumors.) The multiple-sequencing approach is used in most biological sequencing experiment, but not generally in cancer genomics, presumably due to the expense of sequencing multiple replicates for each of the hundreds of samples necessary for cancer genomics research. Nevertheless, even if researchers cannot verify every mutation or sequence multiple replicates for each tumor, it would be useful to know what percentage of called mutations would be likely to appear in additional sequencing replicates.

TCGA's GBM data set includes 2 technical replicates each of 55 tumors. In this case, the technical replicates are not identical. One protocol included an additional amplification step \citep{TCGA-GBM}, and we refer to this replicate as the WGA (whole-genome amplification) replicate, and the other as the WGS (whole-genome sequencing) replicate. Despite the difference in sequencing protocol, we would expect any fully penetrant mutation to appear in both replicates, while incompletely penetrant mutations and sequencing error might appear in only one. Consequently, those putative mutations appearing in both replicates are more likely to be real somatic variants than those found in only one replicate. We further hypothesized that the WGA samples would have a greater number of amplification errors, and thus more putative SNPs per sample, than the corresponding WGS samples.

For each patient (n=55), we called mutations in both technical replicates and in the patient's blood sample, using the same computational pipeline (see Figure~\ref{fig:pipeline} and Methods): We downloaded TCGA BAM files with CGHub \citep{CGHub}, re-generated fastq files with picard \citep{picard}, re-aligned the fastq files to hg19 with bwa \citep{bwa}, performed indel re-alignment and base recalibration with GATK \citep{GATK}, and finally called somatic mutations with SomaticSniper \citep{SomaticSniper}. We then compared the VCFs produced by SomaticSniper for each of the two technical replicates, and calculated the number of somatic mutations called in each replicate and the number of individual somatic mutations called in both replicates (hereafter, the overlap, see Figure~\ref{fig:pipeline}). We further calculated the percentage of mutations in each replicate that occurred in the overlap between the two.

\textbf{copy edit from here}
There were on average 844 putative mutations in the WGS replicates and 1,531 putative mutations in the WGA replicates (for more detail, see summary statistics, Table 1). Across all samples, the number of mutations in a given WGS replicate was correlated with the number of mutations in it's corresponding WGA replicate, cor=0.51, t=4.3, df=53, p-value=8.35e-05 (Figure~\ref{fig:unfiltered_overlap}). As expected, for each sample the WGA replicate (with the additional amplification step) had slightly more mutations overall, with a slightly smaller percentage appearing in the overlap (Figures~\ref{fig:unfiltered_overlap} and~\ref{fig:unfiltered_total_muts}). We further found that the percent overlap between the two samples, calculated by $\text{WGA} \cap \text{WGS}/\text{WGS}$ for WGS replicates and $\text{WGA} \cap \text{WGS}/\text{WGA}$ for WGA replicates, was fairly consistent, on average 31\% in WGA replicates and 44\% in WGS replicates (see summary statistics, Table~\ref{tab:summary_stats}). As expected, the distribution was narrower and taller in the WGS replicates, because on the whole the WGA samples had more amplification errors than the WGS samples (Figure~\ref{fig:unfiltered_overlap}).

% Experiment: number of mutations (figure 4)
%(a) a brief (one sentence to one paragraph) summary of what you did and why
%(b) a description of what you found carrying out this experiment
Although the percent overlap in WGS and WGA samples remained fairly constant, and the number of possible mutations in WGS and WGA lists was correlated, the exact number of putative mutations varied across samples by orders of magnitude (less than 23 to over 10,966). Different cancers mutate at different rates: some pediatric cancers have very few mutations \citep{RB2hit, pediatric}, while some adult tumors show a mutator phenotype leading to vastly more mutations, usually resulting from errors in DNA repair pathways \citep{mutator}. GBM specifically is thought to have a relatively low mutation rate \citep{Parsons, TCGA-GBM-13}, and while some of our samples had low mutation frequencies in line with this theory (29 out of 110 samples had a mutation frequency within a factor of 2 of the reported 3 mutations per Mbp genome), several samples also had mutation frequencies an order of magnitude greater (24 out of 110 samples had a mutation frequency greater than 30 mutations per Mbp genome). One possible explanation is a degraded DNA sample, or bad data. If this were the case, we would expect the percentage of the overlap between replicates (a measure of data quality) to decrease with the overall number of putative mutations. We found no correlation (cor=0.25, p=.06, not significant) between the number of putative mutations in the WGS replicate and the percentage of those mutations that were in the overlap between replicates (Figure~\ref{fig:unfiltered_total_muts}), though it is possible that with more more samples this could become a weak correlation. Still, that the percentage overlap does not correlate strongly with the number of putative mutations in either replicate suggests that some samples may simply have a higher mutation frequency than others, or indeed than is generally supposed in GBM.

@@

We next looked at the individual effects of each of the five filters that we coded and one of the SomaticSniper filters (Variant Allele Quality, or VAQ). We first looked at the total number of putative SNPs removed by each filter (Figure~\ref{fig:boxplot_number_filtered}). We found that different filters removed different numbers of mutations, and that the lion's share of mutations were removed by the VAQ and LOH filters, which removed on average 309 and 539 putative SNPs, respectively (Table~\ref{tab:summary_stats}). Three other filters, those removing overlap with dbSNP and mutations within a 10 bp window of indels or other SNPs, removed 16, 28, and 46 putative mutations per sample (Figure~\ref{fig:boxplot_number_filtered} and Table~\ref{tab:summary_stats}). The final filter, which removed putative SNPs with less than 10\% coverage of the alternate allele, removed 1 putative SNP on average.

We next asked how many of the mutations removed by each filter were mutations present in the overlap between technical replicates, and how many were in just one sample? Put differently, what percentage of putative SNPs removed by a given filter were in the category more likely to be true positives (overlap), versus the category more likely to be false positives (only one replicate)? To answer this, we graphed the percent of the overlap (per sample) that was removed by each of the six filters (Figure~\ref{fig:boxplot_percent_overlap_filtered}). We found that the three filters removing overlap with dbSNP and putative mutations within 10bp of an indel or another SNP removed, on average, only 3\%, 4\%, or 3\% of the overlap, or almost none, respectively (Table~\ref{tab:summary_stats}). In contrast, the VAQ filter (specific to SomaticSniper) and the LOH filter each removed 53\% and 51\% of the overlap, or about half each, respectively (Figure~\ref{fig:boxplot_percent_overlap_filtered}, Table~\ref{tab:summary_stats}). Thus, our evidence suggests that the filters removing overlap with dbSNP and putative mutations near other putative mutations are preferentially removing false positives, but the filters removing low VAQ and LOH are less discriminatory.

%% this is figures 8 and 9, which need to be added to the document and explained now...
Finally, we asked whether the VAQ and LOH filters, responsible between them for removing most of the overlap, were removing the same or different putative SNPs in each sample. We plotted the percent of the overlap filtered out by the LOH filter against the number of putative mutations in the overlap (Figure~\ref{fig:LOH_all}), and the percent of the overlap filtered out by the VAQ filter against the number of putative mutations in the overlap (Figure~\ref{fig:VAQ_all}). We found opposite trends: the percentage filtered out by VAQ decreased, and the percentage filtered out by LOH increased, with increasing length of overlap. The two graphs are almost, but not quite, perfect inverses. But given that (i) each of the LOH and VAQ filters removes about half of the overlap, and (ii) all or almost all of the overlap is removed every time, this is expected. We concluded that the LOH and VAQ filters were not removing the same putative mutations.

@@

All sequence data came from The Cancer Genome Atlas (TCGA) Research Network's Glioblastoma multiforme (GBM) data set \citep{TCGA-GBM}. We downloaded three BAM files for each of 55 patients using CGHub \citep{CGHub}. For each patient, data consisted of one BAM file taken from blood DNA, and two BAM files from tumor DNA, one for each technical replicate. In each case, the only difference in data collection for the two sets of tumor DNA was whether or not an amplification step was performed prior to building a library \citep{TCGA-GBM}.

We created a pipeline for backend processing of all TCGA BAM files, summarized in Figure~\ref{fig:pipeline}, with commands given in Table~\ref{tab:commands}. The custom python code to connect the pipeline is available in a public 
git repository (\texttt{https://github.com/clauswilke/GBM\_genomics}). Our pipeline first regenerated fastq files (original reads) from the TCGA BAM files, which were aligned to hg18, using picard \citep{picard}. Next, we used BWA \citep{bwa} to align the fastq files to hg19, and samtools \citep{SAMtools} to sort, index, and de-duplicate the new BAM file. We used GATK \citep{GATK} to do indel realignment and base recalibration, according to the standard best practices for genomics data \citep{best-practices}. Finally, we predicted somatic variants with SomaticSniper \citep{SomaticSniper}, with the output given in VCF format.

@@

After generating a VCF file will all of the putative somatic variants for each replicate of each sample, we used custom python code (available in public git repository) to list the putative mutations in each VCF, and to calculate the overlap between technical replicates. We then filtered the lists of putative SNPs according to the eight filters described in Table~\ref{tab:filters}, using a combination of command line options and custom python code (available in a public git repository). The eight filters were enacted as follows:

\begin{itemize}
\item[GATK:] The GATK quality score is automatically generated by GATK; GATK recommends discarding all putative mutations with a quality score below 40, which we did using the command line option \texttt{-q 40} when we called SomaticSniper (for the exact command, see Table 2). We did not at any point consider putative mutations removed by this filter, and did not consider its individual action. 
\item[SS:] The SomaticScore is a similar metric calculated by SomaticSniper. As recommended, we removed from consideration all putative mutations with a SomaticScore below 40 by using the command line option \texttt{-Q 40} when we called SomaticSniper (for the exact command, see Table 2). We did not at any point consider putative mutations removed by this filter, and did not consider its individual action. 
\item[VAQ:] The Variant Allele Quality (VAQ) score calculated by SomaticSniper is a third measure of this type. SomaticSniper recommends discarding putative mutations with a VAQ below 40, which we accomplished using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item[LOH:] It is general (but not universal) practice to disregard Loss of Heterozygosity (LOH) in large scale genomics, because LOH to easily results from sequencing errors. We excluded LOH variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item[10bp-SNP:] It is universal or near universal practice to exclude variants within 10 bp of another putative somatic variant, because clusters of putative mutations often indicate an error in reads or sequence alignment. We excluded 10bp-SNP variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item[10bp-INDEL:] It is universal or near universal practice to exclude variants within 10 bp of a putative indel, because clusters of putative mutations often indicate an error in reads or sequence alignment. We excluded 10bp-INDEL variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item[dbSNP:] It is universal or near universal practice to exclude variants that overlap with dbSNP, because the overlap often indicates an amplification error and not a true somatic variant. We excluded dbSNP variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\item[$<10\%$:]) It is universal or near universal practice to exclude variants when the alternate allele coverage is less than 10\%, because the low coverage of the alternate allele often indicates sequencing error. We excluded $<10\%$ variants using a custom python script available in the public git repository. This recommendation is discussed in the Results and Discussion.
\end{itemize}

@@

%% Summary Statistics Table
\begin{landscape}
\begin{table}
\caption{\textbf{TABLE CAPTION} Table explanation.}
\label{tab:summary_stats}
\begin{tabular}{ p{8cm} p{2cm} p{2cm} p{2cm}p{2cm}p{2cm} }
	Quantity & average & median & min & max & stdev \\
	\hline
	No. mutations, WGS & 844 & 328 & 110 & 8,192 & 1,306 \\
	No. mutations, WGA & 1,531 & 694 & 23 & 10,966 & 2,230 \\
	\% mutations in overlap, WGS & 31\% & 31\% & 1\% & 74\% & 20\% \\
	\% mutations in overlap, WGA & 44\% & 45\% & 3\% & 71\% & 13\% \\
	No. putative mutations removed, VAQ & 309 & 164 & 17 & 11,439 & 1,372 \\
	No. putative mutations removed, LOH & 539 & 169 & 3 & 8,538 & 1,051 \\
	No. putative mutations removed, 10bp-SNP & 46 & 19 & 0 & 1,515 & 115 \\
	No. putative mutations removed, 10bp-INDEL & 28 & 16 & 0 & 535 & 45 \\
	No. putative mutations removed, dbSNP & 16 & 8 & 0 & 348 & 33 \\
	No. putative mutations removed, $<$10\% & 1 & 1 & 1 & 36 & 3 \\
	\% overlap removed, VAQ & 53\% & 57\% & 0\% & 100\% & 28\% \\
	\% overlap removed, LOH & 51\% & 52\% & 0\% & 99\% & 29\% \\
	\% overlap removed, 10bp-SNP & 3\% & 2\% & 0\% & 20\% & 3\% \\
	\% overlap removed, 10bp-INDEL & 4\% & 4\% & 0\% & 22\% & 4\% \\
	\% overlap removed, dbSNP & 3\% & 2\% & 0\% & 47\% & 7\% \\
\end{tabular}
\end{table}
\end{landscape}
